---
title: "Lab 2 Problems"
author: "Grace Randall"
date: "2024-01-24"
output: pdf_document
---



```{r setup, include=FALSE}
library(ggplot2)
afdat <- read.csv("/home/guest/applied_stats/Applied_stats/Lab_2/AfrPlots.csv", header = TRUE)
```

# Problem 1
```{r Q1}
# a)
boxplot(afdat$MeanGr,
        main = "Mean Growth",
        xlab = "CM",
        horizontal = TRUE
        )
hist(afdat$MeanGr, xlab = 'Mean Growth', main = 'Histogram of Mean Growth')

# b)
mean(afdat$MeanGr, na.rm = TRUE)
median(afdat$MeanGr, na.rm = TRUE)
var(afdat$MeanGr, na.rm = TRUE)
sd(afdat$MeanGr, na.rm = TRUE)
sd(afdat$MeanGr, na.rm = TRUE)/mean(afdat$MeanGr, na.rm = TRUE)
```

b) The mean of mean growth is 2.7cm. The median is 2.6cm. The variance is 0.80 square cm. The standard diviation is 0.89cm. the coefficient of variation is 0.33.

# Problem 2
```{r Q2}
# 5 or less
sum(dbinom(0:5, size = 20, prob = 0.5))
# more than 5
sum(dbinom(6:20, size = 20, prob = 0.5))

n <- 20
k <- 0:20
p <- 0.5
coinflipPD <- (factorial(n)/(factorial(k)*(factorial(n-k))))*(p^k)*(1-p)^(n-k)
# 5 or less
sum(coinflipPD[1:6])
# more than 5
sum(coinflipPD[7:21])

```
# Problem 3
```{r Q3}
#probability of getting more than 17 right
sum(dbinom(17:20, size = 20, prob = 0.25))
```

# Problem 4
```{r Q4}
#a)
l<-4 
x<-9:13
pd<-l^x/factorial(x)*exp(-l)
sum(pd[1:5])

#b)
sum(dpois(x = 9:13, lambda = 4))

#c)
Dist<-dpois(x = 0:13, lambda = 4)
x<-0:13

plot(x, Dist, type="h",xlim = c(-1, 11), ylim = c(0, 0.5),
las = 1, lwd = 2, col = "blue",
ylab = "Probability", xlab = "Number of butterflies")
points(x, Dist, pch = 16, cex = 1.2, col = "dark red")

```

# Question 5
```{r Q5a}

#a)
# Define number of experiments
nexp <- 1000
sample_means_pres <- matrix(NA,nrow=nexp)

# experiment
for (i in 1:nexp){
x <-rbinom(1000,1,0.3)
sample_means_pres[i] <- mean(x)
}
#Plot
sample_means_pres.df <- data.frame(means=sample_means_pres)

pres <- ggplot(sample_means_pres.df,aes(means))+
geom_histogram()+
geom_vline(xintercept=0.3,color="red")

pres

```
the sampling distribution seems to follow a normal distribution.

```{r Q5b}
#b)
# Define number of experiments
nexp <- 1000
sample_means_num <- matrix(NA,nrow=nexp)

# experiment
for (i in 1:nexp){
x <-rpois(1000,5)
sample_means_num[i] <- mean(x)
}
#Plot
sample_means_num.df <- data.frame(means=sample_means_num)

num <- ggplot(sample_means_num.df,aes(means))+
geom_histogram()+
geom_vline(xintercept=5,color="red")

num
```
the sampling distribution seems to follow a normal distribution.

```{r Q5c}
#c)
# Define number of experiments
nexp <- 1000
sample_means_DBH <- matrix(NA,nrow=nexp)

# experiment
for (i in 1:nexp){
x <-rnorm(1000,12,2)
sample_means_DBH[i] <- mean(x)
}
#Plot
sample_means_DBH.df <- data.frame(means=sample_means_DBH)

DBH <- ggplot(sample_means_DBH.df,aes(means))+
geom_histogram()+
geom_vline(xintercept=12,color="red")

DBH
```
the sampling distribution seems to follow a normal distribution.
