---
title: "Lab 1"
author: "Grace Randall"
date: "2024-01-22"
output: pdf_document
---
## Setup 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r data set up}
x <- c(0,1,2,3,4,5,6,7,8,9,10)
y <- c(2.96, 4.20, 2.84, 3.84, 6.57, 6.95, 9.32, 10.57, 9.72, 11.57, 11.53)

tree_data <- data.frame(height=x,diameter=y)

```

## Question 1

```{r Q1}

#a
mean(tree_data$height)

mean(tree_data$diameter)

#b
quantile(tree_data$height,probs=0.5)

quantile(tree_data$diameter,probs=0.5)

```

a) the mean tree height is 5m. The mean tree diameter is 7cm.

b) the median tree height is 5m. The median tree diameter is 7cm.

## Question 2
```{r Q2a}
# a)
range(tree_data$height)

range(tree_data$diameter)

max(tree_data$diameter)-min(tree_data$diameter)

```
a) The range of the tree height is 10m from 0m to 10m. the range of tree diameter is 8.73cm from 2.84cm to 11.57cm.

```{r Q2b}
# b)

var(tree_data$height)

var(tree_data$diameter)

```
b) The variance in tree height is 11 square meters. The variance in tree diameter is 11.77 square centimeters

```{r Q2c}
# c)

sd(tree_data$height)

sd(tree_data$diameter)

```
c) The standard deviation of tree height is 3.31m. The standard deviation of tree diameter is 3.43cm.

## Question 3

```{r Q3a}
# a)

ggplot(tree_data,aes(x = height, y = diameter))+
  geom_point()+
  xlab("Height (m)")+
  ylab("Diameter (cm)")

```

b) I hypothesize that as tree height increases so does as well

c) A linear regression with m=1 and b=2 would mean that as tree height increases by 1m the tree diameter would increase by approximately 1cm and that trees of 0m of height would have a diameter of 2cm.

```{r Q3d}
# d)

ggplot(tree_data,aes(x = height, y = diameter))+
  geom_abline(intercept = 2, slope = 1, color="lightblue",size=1)+
  geom_point()+
  xlab("Height (m)")+
  ylab("Diameter (cm)")

```

